Towards Practical First-Order Model Counting

### 2023-08-30 paper.tex
Some writing and editing

% See https://www.overleaf.com/learn/latex/theorems_and_proofs
% for a nice explanation of how to define new theorems, but keep
% in mind that the amsthm package is already included in this
% template and that you must *not* alter the styling.
\newtheorem{example}{Example}
\newtheorem{theorem}{Theorem}

@@

\DeclareMathOperator{\Dom}{Dom}
\DeclareMathOperator{\Doms}{Doms}
\DeclareMathOperator{\Preds}{Preds}
\DeclareMathOperator{\Vars}{Vars}

@@

\section{Methodology}

\begin{itemize}
  \item Added support for infinite precision integers using the \texttt{gmp}
        library to the numerical evaluation part. Now, all functions take
        arguments (domain sizes) as unsigned integers and return
        \texttt{mpz\_class} objects (defined in the \texttt{gmpxx.h} header).
  \item Converting the recursive equations into a \texttt{C++} program, which
        can then be compiled and executed to obtain numerical values.
  \item The translation of a set of equations($E$) into \texttt{C++} works as
        follows.
        \begin{enumerate}
          \item First, we create a cache for each function in $E$. For a
                function with name \texttt{func}, this cache is called
                \texttt{func\_cache}. This is implemented as a multi-dimensional
                vector containing objects of \texttt{class cahce\_elem} defined
                as shown in the example code. The default initialization of this
                object is to $-1$ which is useful for recognizing unevaluated
                cases.
          \item Next, we create a function definition for each lhs in $E$,
                including all functions and base cases. The signatures of these
                functions is decided as follows - a function call containing
                only variable arguments is named as the function itself, and
                ones with constants in their arguments are suffixed with a
                string that contains \texttt{'x'} at the $i$th place if the
                $i$th argument is variable and the $i$th argument if that
                argument is a constant. For example, $f(x1, x2, x3)$ is declared
                as \texttt{int f(int x1, int x2, int x3);} and $f(1, x2, x3)$ is
                declared as \texttt{int f\_1xx(int x2, int x3);} (the constant
                arguments are removed from the signature).
          \item The rhs of each equation in $E$ is used to define the body of
                the equation corresponding to the lhs of that equation. The
                function body (for a function \texttt{func} corresponding to
                equation $e$) is formed as follows-
                \begin{enumerate}
                  \item First, we check if the evaluation is already present in
                        the cache. If so, then we return the cache element. The
                        cache accesses are done using the \texttt{get\_elem}
                        function(definition given in the example), which resizes
                        the cache if the accessed index is out of range.
                  \item If the element is absent, then we decide if the
                        arguments corresponding to $e$ or one of the functions
                        corresponding to the base cases, based on the value of
                        the arguments. If it corresponds to the base cases, then
                        we directly call the base case function and return its
                        value. Else, we evaluate the value using the rhs, store
                        the evaluated value in the cache and return the
                        evaluated value. Note that in this step, we only call
                        the base case function with one more constant argument
                        that \texttt{func}. For example, \texttt{f0(x, y)} would
                        call \texttt{f0\_0x(y)} if \texttt{x == 0} and
                        \texttt{f0\_x0(x)} if \texttt{y == 0}.
                  \item In order to translate the rhs, we do the following
                        changes-
                        \begin{itemize}
                          \item \verb|a^b| is converted to \texttt{power(a,b)}.
                          \item \texttt{Sum[exp, x, a, b]} = ($\sum_{x=a}^{b}exp$) is converted to -\\
                                \begin{lstlisting}
([y,z,...](){
    int sum{0};
    for(int x{a}; x <= b; x++)
        sum += exp;
    return sum;
})()
\end{lstlisting}
                                where y, z, ... are the free variables present
                                in \texttt{exp}.
                        \end{itemize}
                \end{enumerate}
        \end{enumerate}
  \item Implemented the evaluation of basecases. The evaluation is done by
        simplifying the clauses and then using CRANE to find the basecases.
        First, in the \texttt{SimplifyUsingWolfram} class, while traversing the
        graph to find the equations, we store 2 \texttt{Map} objects -
        \texttt{clause\_func\_map}(which stores the mapping from the function
        names to the formulae, whose model count they represent) and
        \texttt{var\_domain\_map}(which stores the mapping from the variable
        names to the domains whose sizes they represent). Then, a particular
        domain is selected(using the algorithm described in previous reports)
        and the clauses are simplified. Then, CRANE is called on those clauses
        to evaluate the base cases using \texttt{WeightedCNF.SimplifyInWolfram}.
        After that, we change the function names and variable to make it
        consistent with the previous domain to variable mapping, and append
        these basecases to the set of equations.
  \item Ideas for finding which base cases are needed. The goal is to find the
        sufficient set of base cases. We know that if say, on the rhs of all
        equations, the domain size appears as $m-c_1, m-c_2, \dots, m-c_k$, then
        finding $f(0, x_1, x_2, \dots)$, $f(1, x_1, x_2, \dots)$,
        $\dots f(m_0, x_1, x_2, \dots)$ for every function $f$, where
        $m_0 = \textnormal{max}(c_1, c_2, \dots c_k) - 1$ forms a sufficient set
        of base cases. Hence, in order to do the same efficiently, we can take
        that domain for which $m_0$ is the minimum, i.e.
        $argmin(\textnormal{max}(c_1, c_2, \dots c_k))$.
  \item If a domain is set to $0$, the model count can be found by simplifying
        the universal and existential quantifiers over that domain, and then
        using crane. However, if a domain is set to contain some non-empty
        number of elements, we can apply GDR over that domain and proceed with
        crane, substituting the known values of previously calculated model
        counts in the process.
  \item Implemented a function to simplify functions (recursive and
        non-recursive) using the Wolfram Engine. Modified the wolfram
        simplification functions to incorporate binomial and indicator
        functions.
  \item The report has some (long) examples of formulas being transformed into
        programs perhaps suitable for supplementary material.
\end{itemize}

\subsection{Foo}

@@

The problem with the first approach is that while generating base cases, the
subsequent calls for \textsc{Crane} do not necessarily have the same meanings
for the function arguments and the functions. For example, if \texttt{f1[x0, x1,
  x2]} represents the model count of a constrained formula F, where f1 is an
auxilliary formula and $x0 = |A|, x1 = |B|, x2 = |C|$, ($A,B,C$ are domains) and
we want to evaluate \texttt{f1[0, x1, x2]}($|A| = 0$), then \textsc{Crane} may
return the required model count as \texttt{f0[x0, x1]}, where
$x0 = |B|, x1 = |C|$. We'll need to translate this to \texttt{f1[0,x1,x2]},
which has already been implemented in \texttt{Basecases.scala}. Also, the second
approach can be done in linear time in the length of the formula using
the
\href{https://en.wikipedia.org/wiki/Shunting_yard_algorithm#:~:text=In%20computer%20science%2C%20the%20shunting,abstract%20syntax%20tree%20(AST).}{Shunting
  Yard Algorithm}. Hence, we are implementing the second approach (have already
implemented the parser).

The previous method of base case evaluation on setting a domain to size 0 or 1
had the following error - in case a domain size was set to 0, it assumed that
those predicates which were deleted from the clauses could take any truth value
over the entire domain and the rest were fully covered by the remaining clauses.

For example, consider the constrained CNF formula -
\begin{align}
  \forall x \in \Delta \forall y, z \in \Gamma &: P(x) \lor Q(y, z) \\
  \forall y \in \Gamma^\top &: Q(y, z)
\end{align}
In this case, if we set $|\Delta|$ to $0$, then based on the previous idea, the
new set of clauses would be
\begin{equation}
  \forall y, z \in \Gamma^\top : Q(y, z)
\end{equation}
and the set of removed predicates is $\phi$, and hence the model count returned
would be $1$. However, the actual model count shoud be
$2^{|\Gamma|^2 - |\Gamma^\top|^2}$.

There are 2 ideas which we could think of to solve this -
\begin{enumerate}
  \item \textit{(Currently, we have implemented this)} Convert the clauses
        having universal quantification over the null domain to tautology
        clauses, hence retaining all the predicates which don't have an argument
        belonging to the null domain. For example, we would convert the above
        mentioned CNF formula to
        \begin{align}
          \forall y, z \in \Gamma &: Q(y, z) \lor \neg Q(y, z) \\
          \forall y, z \in \Gamma^\top &: Q(y, z)
        \end{align}
        The model count returned by this will also consider the truth value of Q
        over $y \not \in \Gamma^\top$ or $z \not \in \Gamma^\top$.
  \item We find the domain size over which the predicate has not been defined by
        the clauses and multiply the required factor. For example, in the above
        example, the multiplier would be $2^{|\Gamma|^2 - |\Gamma^\top|^2}$ and
        the simplified set of clauses would be
        \begin{equation}
          \forall y, z \in \Gamma^\top : Q(y, z)
        \end{equation}
        which has a model count of $1$. Hence, the obtained value of model count
        would be $2^{|\Gamma|^2 - |\Gamma^\top|^2} \times 1$.
\end{enumerate}

\section{Algorithms}

For any dependency key, element signature, or clause $P$, argument or variable
$x$, and number or constant $t$, we shall write $P[t / x]$ for the result of
substituting $t$ for all occurrences of $x$ in $P$.

\subsection{Algorithm to Find a Sufficient Set of Base Cases}

\begin{algorithm}[t]
  \caption{Algorithm For Sufficient Base Cases}\label{alg1}
  \KwIn{dependencies}
  \KwOut{base\_cases}

  base\_cases $\gets \emptyset$\;
  \ForEach{dependency $\in$ dependencies}{
    \ForEach{elem $\in$ dependency.value}{
      \ForEach{arg $\in$ elem.args}{
        \For{l $\in$ 0 to arg.const\_subtracted - 1}{\label{line:lim}
          base\_cases += dependency.key[l / arg]\;\label{line:insert}
          \If{dependency.key.func\_name $\ne$ elem.func\_name}{
            base\_cases += elem.signature[l / arg]\;\label{line:signature}
          }
        }
      }
    }
  }
\end{algorithm}

\begin{itemize}
  \item \cref{line:lim}: eg : the limit is 2 for the arg (x-3)
  \item \cref{line:insert} eg : for dependency.key = f(x,y), arg = x and l = 0,
        add f(0, y) to base\_cases
  \item \cref{line:signature} signature for f(x-1, y-2) is f(x,y)
\end{itemize}

The pseudo-code of the algorithm is given as \cref{alg1}. Here,
\texttt{dependencies} maps the function call on the lhs of each equation to the
set of function calls on the rhs of the equation. For example, for the equations
\begin{equation*}
  f(m, n) = g(m-1, n) + f(m-2, n-1)
\end{equation*}
\begin{equation*}
  g(m, n) = f(m-1, n-2) + g(m-1, n-1)
\end{equation*}
the dependencies are
\begin{equation*}
  f(m,n) \mapsto \{g(m-1,n), f(m-2, n-1)\}
\end{equation*}
\begin{equation*}
  g(m, n) \mapsto \{f(m-1, n-2), g(m-1, n-1)\}
\end{equation*}

\subsection{Algorithm to Transform CNF Based on Domain Sizes}

\begin{algorithm}[t]
  \caption{Transforming Formulas Based on Domain Sizes}\label{alg2}
  \KwIn{formula $\phi$, domain $\Delta$, target domain size $n \in \{\, 0, 1 \,\}$, domain size function $|\cdot|$}
  \KwOut{formula $\phi'$, multiplier $m \in \mathbb{N}$}
  $P^{-} \gets \emptyset$; $P^{+} \gets \emptyset$; $\phi' \gets \emptyset$; m $\gets 1$\;
  \uIf{$|\Delta| = 0$}{
    \ForEach{clause $C \in \phi$}{
      \uIf{$\Delta \in \Doms(C)$}{
        $P^{-} \gets P^{-} \cup \Preds(C)$\;
      }
      \Else{
        $P^{+} \gets P^{+} \cup \Preds(C)$\;
        $\phi' \gets \phi' \cup \{\, C \,\}$\;
      }
    }
    $P^{-} \gets P^{-} \setminus P^{+}$\;
    \ForEach{predicate $P \in P^{-}$}{
      m *= $2^{(pred.domain\_sizes.join("*"))}$\;
    }
  }
  \ElseIf{$|\Delta| = 1$}{
    $c \gets$ a new constant symbol\;
    \ForEach{clause $C \in \phi$}{
      $C' \gets C$\;
      \ForEach{$v \in \Vars(C)$ with $\Dom(v) = \Delta$}{
        $C' \gets C'[c / v]$\;
      }
      $\phi' \gets \phi' \cup \{\, C' \,\}$\;
    }
  }
\end{algorithm}
% TODO: either the target domain size or the domain size function needs to go

The pseudo code is given as \cref{alg2}. We use \cref{alg2} to find the
transformed CNF corresponding to each base case obtained using \cref{alg1} and
call \textsc{Crane} on the CNF to obtain the required base cases.

\subsection{Baz}

The following steps were followed while finding the basecases-
\begin{enumerate}
  \item Expand the summations in each equation - Here we expand the summations
        of the form:
        $\sum_{x=0}^{x1}<\textnormal{something}> \cdot [a \le x < b]$, or
        similar inequalities where x is bounded by constants and $a$ and $b$ are
        constants, by substituting the value of x from $a$ to $b-1$. For
        example, we replace
        $\sum_{x=0}^{x_1} {{x_1}\choose{x}} f(x_1-x)\cdot[0 \le x < 2]$ by
        ${{x_1}\choose{0}}f(x_1) + {{x_1}\choose{1}}f(x_1-1)$.
  \item Next, we find the dependencies of those functions that appear on the rhs
        of any equation. Consider, for example the following set of equations-
        \begin{align}
          f0(m, n) &= f1(m-1, n) + f2(m, n-1)\\
          f1(m, n) &= f1(m-1, n-1) \times f2(m-2, n-1)\\
          f2(m, n) &= 2 \times f1(m-3, n-1)
        \end{align}
        In this case, the dependencies computed are
        \begin{align*}
          f1(m, n) &\rightarrow f1(m-1, n-1), f2(m, n-1)\\
          f2(m, n) &\rightarrow f1(m-3, n-1)
        \end{align*}
  \item Now, based on idea II of the previous report, we find a domain that has
        only terms of the form $x-1$ appearing on the rhs of the dependencies.
        The base cases are then calculated by setting this domain size to 0. For
        the above example, $n$ is the selected domain, and not $m$ since there
        are $m-2$ and $m-3$ terms appearing in the arguments.
\end{enumerate}

\paragraph{Limitations of the current implementation.} Ideally, we should
calculate the base cases based on idea II of the previous report by finding the
basecases upto max$(c1, c2, ...) - 1$. However, we are yet to implement code for
finding the basecases for the non-null domain case.

>>>>

% See https://www.overleaf.com/learn/latex/theorems_and_proofs
% for a nice explanation of how to define new theorems, but keep
% in mind that the amsthm package is already included in this
% template and that you must *not* alter the styling.
\newtheorem{example}{Example}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem*{remark}{Remark}

@@

\DeclareMathOperator{\Dom}{Dom}
\DeclareMathOperator{\Doms}{Doms}
\DeclareMathOperator{\Preds}{Preds}
\DeclareMathOperator{\Vars}{Vars}
\DeclareMathOperator*{\argmin}{argmin}

@@

\paragraph{Contributions.}
\begin{itemize}
  \item Converting the recursive equations into a C++ program, which can then be
        compiled and executed to obtain numerical values (see \cref{sec:cpp}).
  \item Support for infinite precision integers using the GNU Multiple Precision
        Arithmetic Library.
\end{itemize}

\section{Snippets}

\begin{itemize}
  \item The evaluation of base cases is done by simplifying the clauses and then
        using \textsc{Crane} to find the base cases. First, while traversing the
        graph to find the equations, we store two maps objects:
        \texttt{clause\_func\_map} (which stores the mapping from the function
        names to the formulae, whose model count they represent) and
        \texttt{var\_domain\_map} (which stores the mapping from the variable
        names to the domains whose sizes they represent). Then, a particular
        domain is selected (using the algorithm described in previous reports),
        and the clauses are simplified. Then, \textsc{Crane} is called on those
        clauses to evaluate the base cases. After that, we change the function
        names and variable to make it consistent with the previous domain to
        variable mapping, and append these base cases to the set of equations.
  \item Finding a sufficient set of base cases. We know that if, say, on the RHS
        of all equations, the domain size appears as
        $m - c_1, m - c_2, \dots, m - c_k$, then finding
        $f(0, x_1, x_2, \dots)$, $f(1, x_1, x_2, \dots)$,
        $\dots f(m_0, x_1, x_2, \dots)$ for every function $f$, where
        $m_0 = \max(c_1, c_2, \dots c_k) - 1$ forms a sufficient set of base
        cases. Hence, in order to do the same efficiently, we can take that
        domain for which $m_0$ is the minimum, i.e.
        $\argmin(\max(c_1, c_2, \dots c_k))$.
\end{itemize}

The previous method of base case evaluation on setting a domain to size zero or
one had the following error. In case a domain size was set to zero, it assumed
that those predicates which were deleted from the clauses could take any truth
value over the entire domain and the rest were fully covered by the remaining
clauses.

For example, consider the formula
\begin{equation}\label{eq:example}
  \begin{split}
    \forall x \in \Delta \forall y, z \in \Gamma &: P(x) \lor Q(y, z) \\
    \forall y \in \Gamma^\top &: Q(y, z).
  \end{split}
\end{equation}
In this case, if we set $|\Delta|$ to zero, the transformed formula would be
\[
  \forall y, z \in \Gamma^\top : Q(y, z),
\]
and the set of removed predicates is empty, and hence the model count returned
would be $1$. However, the actual model count should be
$2^{|\Gamma|^2 - |\Gamma^\top|^2}$.

We solve this problem by converting clauses with universal quantifiers over the
empty domain to tautologies, hence retaining all the predicates that have no
argument assigned to the empty domain. For example, we would convert the above
mentioned formula to
\begin{align*}
  \forall y, z \in \Gamma &: Q(y, z) \lor \neg Q(y, z) \\
  \forall y, z \in \Gamma^\top &: Q(y, z)
\end{align*}
The model count returned by this will also consider the truth value of $Q$ over
$y \not \in \Gamma^\top$ or $z \not \in \Gamma^\top$.

\section{Preliminaries}

TODO: Introduce terms: left-hand side (LHS), right-hand side (RHS).

\begin{definition}
  A \emph{function call} is a term of the form
  $f(x_{1} - c_{1}, \dots, x_{n} - c_{n})$ (written $f(\mathbf{x} - \mathbf{c})$
  for short), where $f$ is an $n$-ary function, each $x_{i}$ is a variable, and
  each $c_{i}$ is a non-negative constant.
\end{definition}
% TODO: do I ever use this term for the case when some of the variables don't
% exist?

\begin{definition}
  A \emph{signature} is a term of the form $f(x_{1}, \dots, x_{n})$ (written
  $f(\mathbf{x})$ for short), where $f$ is an $n$-ary function, and each $x_{i}$
  is a variable. The signature of a function call $f(\mathbf{x} - \mathbf{c})$
  is $f(\mathbf{x})$. For example, the signature of $f(x - 1, y - 2)$ is
  $f(x, y)$.
\end{definition}

\begin{remark}
  The LHS of an equation is always a signature.
\end{remark}

For any signature or clause $C$, argument or variable $x$, and number or
constant $t$, we shall write $C[t / x]$ for the result of substituting $t$ for
all occurrences of $x$ in $C$.
% TODO: check if I'm still using this notation in all these cases
% TODO: explain how this works for function calls too

\section{Identifying a Sufficient Set of Base Cases}

\begin{algorithm}[t]
  \caption{Identifying a set of sufficient base cases}\label{alg1}
  \KwIn{set $D$ of dependencies}
  \KwOut{set $B$ of base cases}

  $B \gets \emptyset$\;
  \ForEach{$(f(\mathbf{x}), g(\mathbf{y} - \mathbf{c})) \in D$}{
    \ForEach{$c_{i} \in \mathbf{c}$}{
      \For{$n \gets 0$ \KwTo $c_{i} - 1$}{\label{line:lim}
        $B \gets B \cup \{\, f(\mathbf{x})[x_{i}/n] \,\}$\;\label{line:insert}
        \lIf{$f \ne g$}{$B \gets B \cup \{\, g(\mathbf{y})[y_{i}/n] \,\}$}\label{line:signature}
      }
    }
  }
\end{algorithm}

The following steps were followed while finding the base cases:
\begin{enumerate}
  \item Expand the summations in each equation. Here we expand the summations of
        the form:
        $\sum_{x=0}^{x_{1}}<\textnormal{something}> \cdot [a \le x < b]$, or
        similar inequalities where x is bounded by constants and $a$ and $b$ are
        constants, by substituting the value of $x$ from $a$ to $b-1$. For
        example, we replace
        $\sum_{x=0}^{x_1} \binom{x_{1}}{x} f(x_1 - x) \cdot [0 \le x < 2]$ by
        $\binom{x_{1}}{0} f(x_1) + \binom{x_{1}}{1} f(x_1-1)$.
  \item Next, we find the dependencies of those functions that appear on the
        RHS of any equation. Consider, for example the following set
        of equations:
        \begin{align*}
          f_{0}(m, n) &= f_{1}(m-1, n) + f_{2}(m, n-1)\\
          f_{1}(m, n) &= f_{1}(m-1, n-1) \times f_{2}(m-2, n-1)\\
          f_{2}(m, n) &= 2 \times f_{1}(m-3, n-1)
        \end{align*}
        In this case, the dependencies computed are
        \begin{align*}
          f_{1}(m, n) &\mapsto \{\, f_{1}(m-1, n-1), f_{2}(m, n-1) \,\}\\
          f_{2}(m, n) &\mapsto \{\, f_{1}(m-3, n-1) \,\}
        \end{align*}
  \item Now, we find a domain that has only terms of the form $x-1$ appearing on
        the RHS of the dependencies. The base cases are then calculated by
        setting this domain size to zero. For the above example, $n$ is the
        selected domain, and not $m$ since there are $m-2$ and $m-3$ terms
        appearing in the arguments.
\end{enumerate}

\paragraph{Limitations of the current implementation.} Ideally, we should
calculate the base cases by finding the base cases up to
$\max(c_{1}, c_{2}, \dots) - 1$. However, currently only empty and singleton
domains are supported.

\begin{itemize}
  \item \cref{line:lim}: the limit is 2 for the arg $(x-3)$.
  \item \cref{line:insert} for $f(\mathbf{x}) = f(x_{1}, x_{2})$,
        $\text{arg} = x$, and $n = 0$, add $f(0, x_{2})$ to $B$.
  \item note that on \cref{line:signature} it is the signature and not the
        original function call.
\end{itemize}

The algorithm is described as \cref{alg1}. Here, a dependency is a pair
$(a, b)$, where $a$ is the signature on the LHS of each equation and $b$ is each
function call on the RHS of the equation. For example, for the equations
\begin{align*}
  f(m, n) &= g(m-1, n) + f(m-2, n-1)\\
  g(m, n) &= f(m-1, n-2) + g(m-1, n-1)
\end{align*}
the dependencies are
\begin{multline*}
  (f(m, n), g(m - 1, n)), (f(m, n), f(m - 2, n - 1)),\\
  (g(m, n), f(m - 1, n - 2), g(m - 1, n - 1)).
\end{multline*}

\section{Algorithm to Transform CNF Based on Domain Sizes}

\begin{algorithm}[t]
  \caption{Transforming Formulas Based on Domain Sizes}\label{alg2}
  \KwIn{formula $\phi$, domain $\Delta$, target domain size $n \in \{\, 0, 1 \,\}$, domain size function $|\cdot|$}
  \KwOut{formula $\phi'$}
  $P^{-} \gets \emptyset$; $P^{+} \gets \emptyset$; $\phi' \gets \emptyset$\;
  \uIf{$|\Delta| = 0$}{
    \ForEach{clause $C \in \phi$}{
      \uIf{$\Delta \in \Doms(C)$}{
        $P^{-} \gets P^{-} \cup \Preds(C)$\;
      }
      \Else{
        $P^{+} \gets P^{+} \cup \Preds(C)$\;
        $\phi' \gets \phi' \cup \{\, C \,\}$\;
      }
    }
    $P^{-} \gets P^{-} \setminus P^{+}$\;
  }
  \ElseIf{$|\Delta| = 1$}{
    $c \gets$ a new constant symbol\;
    \ForEach{clause $C \in \phi$}{
      $C' \gets C$\;
      \ForEach{$v \in \Vars(C)$ with $\Dom(v) = \Delta$}{
        $C' \gets C'[c / v]$\;
      }
      $\phi' \gets \phi' \cup \{\, C' \,\}$\;
    }
  }
\end{algorithm}
% TODO: the removed empty domains should be replaced by smoothing/tautological
% constructions

% TODO: either the target domain size or the domain size function needs to go

We use \cref{alg2} to find the transformed formula corresponding to each base
case obtained using \cref{alg1} and call \textsc{Crane} on the formula to obtain
the required base cases.

\section{Generating C++ Code}\label{sec:cpp}

% TODO: The report has some (long) examples of formulas being transformed into
% programs perhaps suitable for supplementary material.

@@

The problem with the first approach is that, while generating base cases, the
subsequent calls to \textsc{Crane} do not necessarily have the same meanings for
the function arguments and the functions. For example, if
$f_{1}(x_{0}, x_{1}, x_{2})$ represents the model count of a constrained formula
$\phi$, where $f_{1}$ is an auxiliary formula and $x_{0} = |A|$, $x_{1} = |B|$,
$x_{2} = |C|$, and $A$, $B$, $C$ are domains and we want to evaluate
$f_{1}(0, x_{1}, x_{2})$ (i.e., set $|A| = 0$), then \textsc{Crane} may return
the required model count as $f_{0}(x_{0}, x_{1})$, where $x_{0} = |B|$ and
$x_{1} = |C|$. We will then need to translate this to $f_{1}(0, x_{1}, x_{2})$.
Also, the second approach can be done in linear time in the length of the
formula using
the
\href{https://en.wikipedia.org/wiki/Shunting_yard_algorithm#:~:text=In%20computer%20science%2C%20the%20shunting,abstract%20syntax%20tree%20(AST).}{Shunting
  Yard Algorithm}. Hence, we stick to the second approach.

\subsection{Our Approach}

The translation of a set $E$ of equations into a C++ program works as follows.

First, we create a cache for each function in $E$. This is implemented as a
multi-dimensional vector containing objects of \texttt{class cache\_elem}
defined as shown in the example code. The default initialization of this object
is to $-1$ which is useful for recognizing unevaluated cases.

Next, we create a function definition for the LHS of each equation in $E$,
including all functions and base cases. The signatures of these functions is
decided as follows. A function call containing only variable arguments is named
as the function itself, and ones with constants in their arguments are suffixed
with a string that contains \texttt{'x'} at the $i$th place if the $i$th
argument is variable and the $i$th argument if that argument is a constant. For
example, $f(x_{1}, x_{2}, x_{3})$ is declared as \texttt{int f(int x1, int x2,
  int x3);} and $f(1, x_{2}, x_{3})$ is declared as \texttt{int f\_1xx(int x2,
  int x3);} (the constant arguments are removed from the signature).

The RHS of each equation in $E$ is used to define the body of the equation
corresponding to the LHS of that equation. The function body (for a function
\texttt{func} corresponding to equation $e$) is formed as follows.
\begin{enumerate}
  \item First, we check if the evaluation is already present in the cache. If
        so, then we return the cache element. The cache accesses are done using
        the \texttt{get\_elem} function (definition given in the example), which
        resizes the cache if the accessed index is out of range.
  \item If the element is absent, then we decide if the arguments corresponding
        to $e$ or one of the functions corresponding to the base cases, based on
        the value of the arguments. If it corresponds to the base cases, then we
        directly call the base case function and return its value. Else, we
        evaluate the value using the RHS, store the evaluated value in the cache
        and return the evaluated value. Note that in this step, we only call the
        base case function with one more constant argument that \texttt{func}.
        For example, \texttt{f0(x, y)} would call \texttt{f0\_0x(y)} if
        \texttt{x == 0} and \texttt{f0\_x0(x)} if \texttt{y == 0}.
  \item In order to translate the RHS, we convert $\sum_{x=a}^{b} \texttt{exp}$
        to
\begin{lstlisting}
([y,z,...](){
    int sum = 0;
    for(int x = a; x <= b; x++)
        sum += exp;
    return sum;
})()
\end{lstlisting}
        where $y, z, \dots$ are the free variables present in \texttt{exp}.
\end{enumerate}
